---
title: "Descriptive Statistical Analysis of the highest scoring IPL match"
output:
  pdf_document:
    toc: true
date: "`r Sys.Date()`"
---

```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
# CLEAR ENVIRONMENT.
rm(list = ls(all.names = TRUE))

# LOAD PACKAGES

# GLOBAL CHUNK OPTIONS
# 'dev' checks if it's PDF to use high-quality rendering, otherwise standard
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  fig.width = 6,
  fig.height = 4
)

# GGPLOT THEME
theme_set(theme_minimal())  # check ?theme_bw for other themes available
```

\newpage

# Introduction

This report presents a descriptive analysis of the highest-scoring IPL match between **Sunrisers
Hyderabad** (SRH) and **Royal Challengers Bengaluru** (RCB) (15 April 2024), using ball-by-ball data. Each observation corresponds to a single delivery.

# Data Loading and Preprocessing
```{r data-loading-and-processing}
ipl_data <- read.csv("data/IPL.csv") # Load the IPL data. We do not need external packages for reading `CSV`.

# Aggregate "Current Run Rate" variable.
ipl_data$current_run_rate <- round(ipl_data$team_runs / ipl_data$ball_no, 2)

included_columns <- c("batter", "bowler", "batter_balls", "batter_runs", "batting_team",
                      "current_run_rate", "over", "runs_total", "team_runs")

# We will work with a single match i.e., SRH vs RCB, 15th April 2024.
match <- subset(x = ipl_data, subset = match_id == 1426268, select = included_columns)
```

## Data Summary
```{r data-summary}
head(match)
cat("Dataset dimensions:", nrow(match), "rows x", ncol(match), "columns")
```
# Univariate Analysis
## Batter - Absolute frequency of balls faced.
