---
title: "IBM HR Employee Attrition Analysis"
subtitle: "Statistical Computing Project"
author: "Amaan Shaikh | Danish Ali Shaikh | Rajat Pandey"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    toc_depth: 2
    number_sections: true
    fig_caption: yes
geometry: "left=0.75in, right=0.75in, top=0.8in, bottom=0.8in"
fontsize: 11pt
---

```{r setup, include=FALSE}
# CLEAR ENVIRONMENT
rm(list = ls(all.names = TRUE))

# GLOBAL CHUNK OPTIONS
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.width = 7,
  fig.height = 3.5,
  fig.align = "center",
  out.width = "90%"
)
```

# Introduction

The IBM HR Employee Attrition dataset serves as a sample of workforce data, organized as a data matrix of observations and qualitative and quantitative variables. This report performs a comprehensive descriptive statistical analysis to summarize the distribution of individual variables using frequencies, measures of location, and measures of dispersion. Bivariate techniques, including contingency tables and correlation coefficients, are applied to investigate the joint behavior and co-dependency between employee attributes and attrition. Furthermore, simple linear regression is utilized to model the linear relationship between tenure and monthly income by estimating the regression coefficients.

## Dataset & Data Preparation
We analyze ten key variables classified into the following datatypes :

- **Nominal**: `Attrition`, `OverTime`
- **Ordinal**: `JobLevel` (1-5), `JobSatisfaction` (1-4), `WorkLifeBalance` (1-4)
- **Numeric (Discrete)**: `Age`, `YearsAtCompany`, `TotalWorkingYears`, `NumCompaniesWorked`
- **Numeric (Continuous)**: `MonthlyIncome`

```{r load-data, echo=TRUE}
# Load data and select variables
hr <- read.csv("WA_Fn-UseC_-HR-Employee-Attrition.csv", stringsAsFactors = FALSE)
hr_sub <- hr[, c("Attrition", "OverTime", "JobLevel", "JobSatisfaction", 
                 "WorkLifeBalance","Age", "YearsAtCompany", "TotalWorkingYears", 
                 "NumCompaniesWorked", "MonthlyIncome")]

# Convert data types
hr_sub$Attrition <- factor(hr_sub$Attrition)
hr_sub$OverTime  <- factor(hr_sub$OverTime)
hr_sub$JobLevel <- factor(hr_sub$JobLevel, levels = 1:5, ordered = TRUE)
hr_sub$JobSatisfaction <- factor(hr_sub$JobSatisfaction, levels = 1:4, ordered = TRUE)
hr_sub$WorkLifeBalance <- factor(hr_sub$WorkLifeBalance, levels = 1:4, ordered = TRUE)
```

## Data Summary

```{r data-summary}
# Dataset dimensions
dim_table <- data.frame(
  Rows = nrow(hr_sub),
  Columns = ncol(hr_sub)
)

knitr::kable(dim_table, caption = "Dataset Dimensions")

# Summary statistics
summary(hr_sub)

# First 6 rows
knitr::kable(
  t(head(hr_sub)),
  caption = "First 6 Observations (Transposed View)",
  booktabs = TRUE
)
```

# Univariate Analysis

Univariate analysis examines each variable individually to understand its distribution, measures of location & measures of dispersion.

## Categorical and Ordinal Variables

### Attrition Status

Here we tabulate the absolute and relative frequencies of employees who left the company versus those who stayed.
```{r attrition-freq}
# Frequency and proportion table
attrition_tab <- as.data.frame(table(hr_sub$Attrition))
colnames(attrition_tab) <- c("Attrition", "Frequency")

attrition_tab$Proportion <- round(
  attrition_tab$Frequency / sum(attrition_tab$Frequency), 3
)

knitr::kable(
  attrition_tab,
  caption = "Attrition Status: Frequency and Proportion",
  align = c("l", "c", "c"),
  format = "markdown"
)
```

### Overtime Status
This shows the distribution of employees working overtime versus standard hours.
```{r overtime-freq}
# Frequency and proportion table
overtime_tab <- as.data.frame(table(hr_sub$OverTime))
colnames(overtime_tab) <- c("OverTime", "Frequency")

overtime_tab$Proportion <- round(
  overtime_tab$Frequency / sum(overtime_tab$Frequency), 3
)

knitr::kable(
  overtime_tab,
  caption = "Overtime Status: Frequency and Proportion",
  align = c("l", "c", "c"),
  format = "markdown"
)
```

### 3.1.3 Job Level (Ordinal)

The frequency distributions reveal employee composition. The attrition rate indicates what proportion of employees left the company. Job level cumulative frequencies show the hierarchical distribution from entry-level to senior positions.

```{r joblevel-freq}
# Frequency table with percentages and cumulative frequencies
joblevel_freq <- as.data.frame(table(hr_sub$JobLevel))
colnames(joblevel_freq) <- c("Job Level", "Frequency")
joblevel_freq$Percentage <- round(100 * joblevel_freq$Frequency / sum(joblevel_freq$Frequency), 2)
joblevel_freq$`Cumulative Frequency` <- cumsum(joblevel_freq$Frequency)
joblevel_freq$`Cumulative %` <- cumsum(joblevel_freq$Percentage)

knitr::kable(joblevel_freq,
             caption = "Job Level Distribution (Ordinal)",
             align = c('c', 'c', 'c', 'c', 'c'),
             format = "markdown")
```
