---
title: "IBM HR Employee Attrition Analysis"
subtitle: "Statistical Computing Project"
author: "Amaan Shaikh | Danish Ali Shaikh | Rajat Pandey"
date: "`r Sys.Date()`"
output:
  pdf_document:
geometry: "left=0.75in, right=0.75in, top=0.8in, bottom=0.8in"
fontsize: 11pt
---

## 1. Introduction

This document provides a short documentation of the IBM HR Employee Attrition dataset used for the Statistical Computing group project. Instead of just a script to load the data we thought of sharing a summary so that it is easier for you to assess the dataset. 

The dataset was obtained from Kaggle and was originally created by IBM to study employee attrition patterns within an organization.

**Source:** IBM HR Analytics Employee Attrition & Performance  
**URL:** https://www.kaggle.com/datasets/pavansubhasht/ibm-hr-analytics-attrition-dataset

The purpose of this documentation is to summarize the dataset, describe the selected variables and their data types.

## 2. Dataset Description

This is a fictional data set created by IBM data scientists. It includes demographic information, job-related attributes, compensation details, and work–life balance indicators.

The primary target variable is **Attrition**, which indicates whether an employee has left the company.  
Since the original dataset contains many variables, we focus on a subset of variables that are most relevant for attrition analysis and suitable for statistical exploration.

```{r load-data}
# Load dataset.
hr <- read.csv("WA_Fn-UseC_-HR-Employee-Attrition.csv", stringsAsFactors = FALSE)

# Our selection of variables of interest.
data_selected <- hr[, c("Attrition", "OverTime", "JobLevel", "JobSatisfaction", 
                        "WorkLifeBalance", "Age", "YearsAtCompany", "TotalWorkingYears", 
                        "NumCompaniesWorked", "MonthlyIncome")]
```

\newpage

## 3. Summary of the Dataset

### 3.1 Number of Rows and Columns

```{r dimensions}
dim(data_selected)
```

### 3.2 Summary of Variables

```{r summary}
summary(data_selected)
```

## 4. Variable Description

| Variable Name | Data Type | Scale of Measure | Description |
|---------------|-----------|------------------|-------------|
| **Attrition** | Binary | Nominal | Whether the employee left the company (Yes / No) |
| **OverTime** | Binary | Nominal | Whether the employee works overtime |
| **JobLevel** | Categorical | Ordinal | Job level on a scale from 1 (lowest) to 5 (highest) |
| **JobSatisfaction** | Categorical | Ordinal | Job satisfaction level (1 = Low, 4 = Very High) |
| **WorkLifeBalance** | Categorical | Ordinal | Work-life balance rating (1 = Bad, 4 = Best) |
| **Age** | Numeric | Discrete | Age of the employee (in years) |
| **YearsAtCompany** | Numeric | Discrete | Number of years the employee has worked at the company |
| **TotalWorkingYears** | Numeric | Discrete | Total number of years of professional experience |
| **NumCompaniesWorked** | Numeric | Discrete | Number of companies the employee has previously worked for |
| **MonthlyIncome** | Numeric | Continuous | Monthly salary of the employee (in USD) |

## 5. Example Data (First Few Observations)

```{r head-data}
head(data_selected, 5)
```

## 6. Conclusion

We found the IBM HR employee attrition dataset suitable for the given task as it has simple yet sufficient variables to perform descriptive statistical analysis.